Order reduction and MILP models for real-time DNN robustness certification.

### 2025-08-02 AAAI2026/abstract.tex
title!!

In this paper, we consider {\em global} robustness, that is, guarantees not restricted to a set of local images. We focus on deriving {\em bounds} on the variation of output values across different decision classes of a DNN under a given perturbation. This global verification problem is significantly more complex than local robustness, as the number of variables doubles (from the deviation image to the image and its deviation). Further, the values each neuron can take are no longer limited to a small neighborhood. To obtain usable bounds, we develop several novel partial MILP models for global robustness, with different trade-offs. Last, we use order reduction techniques to reduce the space of inputs considered, avoiding unrealistic images, by using principal component analysis (PCA). This results into the real-time  certification of the robustness for $86\%$ of incoming images in the MNIST benchmark for an $L_1$-perturbation of 0.5, as well as for a surrogate computing the
%latent
plastic strain associated with a pipe deformation map.

>>>>

In this paper, we consider {\em global} robustness, that is, guarantees not restricted to a set of local images. We focus on deriving {\em bounds} on the variation of output values across different decision classes of a DNN under a given perturbation. This global verification problem is significantly more complex than local robustness, as the number of variables doubles (from the deviation image to the image and its deviation). Further, the values each neuron can take are no longer limited to a small neighborhood. To obtain usable bounds, we develop several novel  MILP models for global robustness, with different trade-offs. Last, we use order reduction techniques to reduce the space of inputs considered, avoiding unrealistic images, by using principal component analysis (PCA). This results into the {\em real-time} certification of the robustness for $86\%$ of incoming images in the MNIST benchmark for an $L_1$-perturbation of $.5$, as well as for a surrogate computing the 
%latent
plastic strain associated with a pipe deformation map.

### 2025-08-02 AAAI2026/experiments.tex
title!!

\begin{figure*}[t!]
	\centering
\includegraphics[scale=0.6]{image.png} \hspace{1.5cm}
\includegraphics[scale=0.6]{perturb.png}
\caption{An improbable image and its perturbation ($L1$-difference of $.5$ in a unique pixel at x=14, y=8 from top) with maximal $\beta^{.5}_{6,8}=0.518$ for MNIST as obtained by the "3v" $500 \times 2$ model in {\bf full 784 dimension} image space.}
\label{fig3}
\end{figure*}

\begin{figure*}[t!]
	\centering
\includegraphics[scale=0.5]{redimage.png} \hspace{1.5cm}
\includegraphics[scale=0.5]{redperturb.png}
\caption{An image and its perturbation from the {\bf 20-dimension reduced space} with maximal $\beta^{.5}_{6,8}=.084$ for MNIST as obtained by the "2v" $500 \times 2$ model.}
\label{fig4}
\end{figure*}

>>>>

\begin{figure*}[t!]
	\centering
\includegraphics[scale=0.6]{image.png} \hspace{1.5cm}
\includegraphics[scale=0.6]{perturb.png}
\caption{An improbable image for MNIST and its perturbation ($L1$-difference of $.5$ in a unique pixel at x=14, y=8 from top) with maximal $\beta^{.5}_{6,8}=0.518$, as obtained by the "3v" $500 \times 2$ model in {\bf full 784 dimension} image space.}
\label{fig3}
\end{figure*}

\begin{figure*}[t!]
	\centering
\includegraphics[scale=0.5]{redimage.png} \hspace{1.5cm}
\includegraphics[scale=0.5]{redperturb.png}
\caption{An image for MNIST and its perturbation from the {\bf 20-dimension reduced space} with maximal $\beta^{.5}_{6,8}=.084$, as obtained by the "2v" $500 \times 2$ model.}
\label{fig4}
\end{figure*}

### 2025-08-02 AAAI2026/introduction.tex
title!!

The benchmarks usually focus on {\em local} robustness, i.e. given a DNN, an image and a small neighborhood around this image, is it the case that all the images in the neighborhood are classified in the same way by the DNN? The neighborhood is provided by a maximal perturbation of the input image, often an $L_\infty$ perturbation, i.e. every subpixel of the input image can vary in a very small range, typically $\frac{2}{255}$ (that is 2 levels of grey/blue/right/green), as $L_\infty$ is perfectly linear and perturbations of subpixels are independent. While the $L_\infty$ norm is the easiest to handle, it is not necessarily the most meaningful perturbation. More importantly, verification tools for local robustness are hard to be used in a real-time decision making pipeline: considering an autonomous car with a video feed from the dashboard, it is computationally too intensive to certify in few ms images of the feed to e.g. skip non-robust images and only consider certified robust images for the decision-making process.

@@

	\item Results obtained on the full input space have several shortcomings: while the bounds are fully correct, they are also particularly pessimistic, as all inputs, including absolutely improbable ones very far away from anything the network has been trained on, need to be accounted for. Also, the runtime to obtain the bound is particularly long. 
	%Finally, when computing worst-case 
	%pairs (image, perturbation), improbable images are generated, hence these  are not meaningful. 
	To solve this problem, we restrict the input space to ignore improbable inputs, using model order reduction techniques from engineering science \cite{aiware}. Specifically, we use Principal Component Analysis (PCA) to reduce the space to a linear input space. We choose the number of dimensions of the space to equal the accuracy of the DNN on the reduced space. 
	%(using a projection to the reduced space then the inverse projection to obtain a very similar image understandable by the DNN). 
	On the MNIST benchmark, this means 20 linear dimensions to match the $97\%$ accuracy of the DNN we considered, instead of the 784 dimensions of the full image space. Using PCA, which is linear, makes it easy to specify perturbations on the actual image (where it is meaningful) rather than on the reduced space.

\item Experimentally, this allows to obtain bounds for a considered pre-trained DNN for MNIST (with accuracy $97\%$) that can certify in real time that $86\%$ of the images are robust for an $L_1$ perturbation of $.5$, resp. $53\%$ for an $L_1$ perturbation of $1$. We finally experimented our MILP models for a {\em regression} task as well, considering a DNN surrogate built from a slow physical finite element model, providing the deformation and the plastic strain of a pipe \cite{aiware}. The DNN has been learnt on a reduced order basis already of dimension 10, from thousands of input dimensions. For regression, the aim is to understand if the DNN surrogate that has been learnt is brittle to small perturbations, which could make it unfaithful to the physical model. Here we want to obtain bounds (which is what our models provide), as well as generating worst-case (image,perturbation) pair, where the strain is particularly different, close to the bound, and that can be run in the physical model to understand whether the DNN surrogate is faithful. Our tool was helpful to provide limited bounds, as well as an almost matching worst-case (image,perturbation) pair that was confirmed in the physical model.
Overall, the "2v" model generates accurate bounds. The decoupling of variables generating the "3v" model offers good trade-offs, reaching slightly lower bounds in the same runtime. 
The "1v" model is faster to converge, but also generates larger bounds, unless runtime is very restricted (Table \ref{table.pipe}).
\end{enumerate}

>>>>

The benchmarks usually focus on {\em local} robustness, i.e. given a DNN, an image and a small neighborhood around this image, is it the case that all the images in the neighborhood are classified in the same way by the DNN? The neighborhood is provided by a maximal perturbation of the input image, often an 
$L_\infty$-perturbation, i.e. every subpixel of the input image can vary in a very small range, typically $\frac{2}{255}$ (that is 2 levels of grey/blue/right/green), as $L_\infty$ is perfectly linear and perturbations of subpixels are independent. While the $L_\infty$-norm is the easiest to handle, it is not necessarily the most meaningful perturbation. More importantly, verification tools for local robustness are too computationally-intensive to be used in a real-time decision making pipeline: considering an autonomous car with a video feed from the dashboard, 
images of the feed cannot be certified robust in few ms to e.g. skip non-robust images and only consider certified robust images for the decision-making process.

@@

	\item Results obtained on the full input space have several shortcomings: while the bounds are fully correct, they are also particularly pessimistic, as all inputs, including absolutely improbable ones very far away from anything the network has been trained on, need to be accounted for. Also, the runtime to obtain the bound is particularly long. 
	%Finally, when computing worst-case 
	%pairs (image, perturbation), improbable images are generated, hence these  are not meaningful. 
	To solve this problem, we restrict the input space to ignore improbable inputs, using model order reduction techniques from engineering science \cite{Paco}. Specifically, we use Principal Component Analysis (PCA) to reduce the space to a linear input space. We choose the number of dimensions of the space to equal the accuracy of the DNN on the reduced space. 
	%(using a projection to the reduced space then the inverse projection to obtain a very similar image understandable by the DNN). 
	On the MNIST benchmark, this means 20 linear dimensions to match the $97\%$ accuracy of the DNN we considered, instead of the 784 dimensions of the full image space. Using PCA, which is linear, makes it easy to specify perturbations on the actual image (where it is meaningful) rather than on the reduced space.

\item Experimentally, this allows to obtain bounds for a considered pre-trained DNN for MNIST (with accuracy $97\%$) that can certify in {\em real-time} that $86\%$ of the images are robust for an $L_1$ perturbation of $.5$, resp. $53\%$ for an $L_1$ perturbation of $1$. We finally experimented our MILP models for a {\em regression} task as well, considering a DNN surrogate built from a slow physical finite element model, providing the deformation and the plastic strain of a pipe \cite{aiware}. The DNN has been learnt on a reduced order basis already of dimension 10, from thousands of input dimensions. The aim is to understand if the DNN surrogate that has been learnt is brittle to small perturbations, which could make it unfaithful to the physical model. Here we want to obtain bounds (which is what our models provide), as well as generating worst-case (image,perturbation) pair, where the strain is particularly different, close to the bound, and that can be run in the physical model to understand whether the DNN surrogate is faithful. Our tool was helpful to provide limited bounds, as well as an almost matching worst-case (image,perturbation) pair that was confirmed in the physical model.
Overall, the "2v" model generates accurate bounds. The decoupling of variables generating the "3v" model offers good trade-offs, reaching slightly lower bounds in the same runtime. 
The "1v" model is faster to converge, but also generates larger bounds, unless runtime is very restricted (Table \ref{table.pipe}).
\end{enumerate}
